PAPER DRAFT

### 2021-11-10 Papers/Polymethylation_scores/05_results.Rmd

Of `r methylCohort$MethID%>%length()` samples from `r methylCohort$idnum%>%unique()%>%length()` unique participants with DNA methylation data measured on the Illumina 450K  array, `r m_obs` samples from `r n_total` participants had complete data on additional key covariates (Figure 1). There were `r n_twovisit` participants with both age 9 and age 15 DNA methylation samples. Excluded samples were similar to included samples, except that included samples were slightly more likely to be from children of African genetic ancestry and less likely to be from children of Hispanic genetic ancestry (Supplemental Table 1). In the included sample, 20% percent of the mothers reported any prenatal maternal smoking, `r round(table(modeldata_wide$m1g2_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal alcohol use and `r round(table(modeldata_wide$m1g3_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal drug use. The mean income to poverty ratio of mothers at birth was `r round(mean(modeldata_wide$cm1inpov), 2)`. Of the children in the analytic sample, `r round(table(modeldata_wide$cm1bsex)/nrow(modeldata_wide)*100)['1 Boy']`% were male, `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['African ancestry']`% were of African genetic ancestry, and `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['Hispanic ancestry']`% were of Hispanic genetic ancestry.

@@

In bivariate analyses and multivariable models, we focused on the polymethylation scores and the single top CpG site from Joubert et al, AHRR:cg05575921, as hypothesized biomarkers, and used global methylation and the pediatric clock as negative controls.

@@

Mothers who reported smoking during pregnancy had lower income to poverty ratios than those who did not ( `r modeldata_people%>%filter(smkPreg_binary=='Yes')%>%summarise(mean(cm1inpov))%>%round(2)` vs `r modeldata_people%>%filter(smkPreg_binary=='No')%>%summarise(mean(cm1inpov))%>%round(2)`).  Mothers who smoked were more likely to report prenatal alcohol use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%), prenatal drug use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%) and postnatal smoking (`r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[1, 2]%>%round(2)*100`%) than mothers who did not report prenatal maternal smoking (Table 1).  Children of mothers who reported smoking during pregnancy were more likely to be of European genetic ancestry (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[2, 2], 2)*100`%) than children of mothers who did not (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[1, 2], 2)*100`%, *P* value=`r round(with(modeldata_people, chisq.test(table(smkPreg_binary, ancestry)))$p.value, 4)`). Children of mothers who reported smoking during pregnancy had higher prenatal maternal smoking polymethylation scores than children of mothers who did not at both the age 9 (`r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) and age 15 (`r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) visits (Table 1, Supplemental Table2). At age 9, children exposed to prenatal maternal smoking had lower DNA methylation at AHRR:cg05575921 (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)*100`%) than children of non-smoking mothers (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)*100`%, *P* value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`). At age 15, children exposed to prenatal smoking had lower DNA methylation at AHRR:cg05575921 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)*100`%) than children of non-smoking mothers, although this difference was not significant (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)*100`% *P* value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).

@@

The association between prenatal maternal smoking and the polymethylation score was also observed in multivariable models  (Figure 3; Supplemental Table 2). At age 9, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score.  At age 15, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score. A consistent association was observed when stratifying to the African genetic ancestry sample (n= `r modeldata_people%>%filter(ancestry=='African ancestry')%>%nrow()`). In the African genetic ancestry sample at age 9, prenatal maternal smoking was associated with a `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score. The direction and magnitude of the association was similar in the European and Hispanic genetic ancestry samples at age 9, though attenuated at age 15 (Figure 3; Supplemental Tables 2).

Prenatal maternal smoking was associated with lower methylation at AHRR:cg055975921 at age 9 (`r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921'", 'smkPreg_binaryYes', 2)`) and age 15 (`r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921'", 'smkPreg_binaryYes', 2)`). Associations between prenatal maternal smoking and AHRR:cg05575921 were less robust across ancestries (Figure 3).

## Sensitivity analyses

When also adjusting for other prenatal exposures and postnatal smoke exposure, the association between prenatal maternal smoking and the polymethylation score was very consistent (Supplemental Figure 4). The direction of the association between prenatal maternal smoking and AHRR:cg055975921 was also consistent, although no longer significant at age 15. Children exposed to postnatal smoke but not prenatal smoke did not have higher polymethylation scores than children unexposed to both postnatal and prenatal smoke.  (Supplemental Figure 5). Results were also similar when controlling for surrogate variables instead of known covariates (Supplemental Figure 6; Supplemental Table 2).

Results were similar when using linear mixed effect models instead of age-stratified models (Supplemental Table 3 & 4).

Results were also similar when using polymethylation scores constructed using alternative regression coefficients (see Methods, Supplemental Figure 7). The association between prenatal maternal smoking and the polymethylation score using regression coefficients from newborn cord blood with cell-type controls was the strongest.  Prenatal maternal smoking also associated with other selected *a priori* CpG sites consistently with associations reported in previous meta-analyses (Supplemental Table 2, Supplemental Figure 8).

@@

Next, we compared the accuracy of different DNA methylation summary measures for classifying prenatal maternal smoking using receiver operating curves (Figure 4A; Supplemental Table 5). We considered classification when using DNA methylation summary measures alone. We also considered classification when using DNA methylation summary measures in addition to a base model of child sex, maternal income to poverty ratio at baseline, child age at methylation measurement, estimated immune cell proportion, plate from methylation processing and the first two components of genetic ancestry. When used without DNA methylation measures, these base model variables had an area under the curve (AUC) of `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 9')%>%pull(auc)` at age 9 and `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 15')%>%pull(auc)` at age 15.

At age 9, including the polymethylation score  significantly improved the base model accuracy (AUC: `r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(auc)`, *P* value comparing to base model=`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(4)`). Similarly, including the polymethylation significantly improved the base model accuracy at age 15 (AUC:`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(auc)` , *P* value=`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(4)`).

At age 9,the base model with the polymethylation score had a larger AUC than the base model with AHRR:cg05575921 (AUC base model+AHRR:cg05575921: `r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(auc)`; *P* value =`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(5)`). At age 15, this was also true (AUC for base model+AHRR:cg05575921=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(auc)`,  *P* value=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(5)`). Accuracy was not improved by including polymethylation scores from age 9 and age 15 together (Figure 4B).

## Sensitivity analyses

>>>>

In a sample of the Fragile Families and Child Wellbeing Study, saliva DNA methylation was measured on `r methylCohort$MethID%>%length()` samples from `r methylCohort$idnum%>%unique()%>%length()` unique participants with the Illumina 450K  array. Complete data on covariates of interest was available on `r m_obs` samples from `r n_total` participants who were included in the analysis (Figure 1). There were `r n_twovisit` participants with both age 9 and age 15 DNA methylation samples. Excluded samples were similar to included samples, except that included samples were slightly more likely to be from children of African genetic ancestry and less likely to be from children of Hispanic genetic ancestry (Supplemental Table 1). In the included sample, 20% percent of the mothers reported any prenatal maternal smoking, `r round(table(modeldata_wide$m1g2_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal alcohol use and `r round(table(modeldata_wide$m1g3_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal drug use. The mean income to poverty ratio of mothers at birth was `r round(mean(modeldata_wide$cm1inpov), 2)`. Of the children in the analytic sample, `r round(table(modeldata_wide$cm1bsex)/nrow(modeldata_wide)*100)['1 Boy']`% were male, `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['African ancestry']`% were of African genetic ancestry, and `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['Hispanic ancestry']`% were of Hispanic genetic ancestry.

@@

In bivariate analyses and multivariable models, we focused on the polymethylation scores and the single top CpG site from prior research, cg05575921 in the *AHRR* gene,, as hypothesized biomarkers, and used global methylation and the pediatric clock as negative controls.

@@

Mothers who reported smoking during pregnancy had lower income to poverty ratios (`r modeldata_people%>%filter(smkPreg_binary=='Yes')%>%summarise(mean(cm1inpov))%>%round(2)`) than those who did not ( `r modeldata_people%>%filter(smkPreg_binary=='No')%>%summarise(mean(cm1inpov))%>%round(2)`).  Mothers who smoked were more likely to report prenatal alcohol use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%), prenatal drug use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%) and postnatal smoking (`r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[1, 2]%>%round(2)*100`%) than mothers who did not report prenatal maternal smoking (Table 1).  Children of mothers who reported smoking during pregnancy were more likely to be of European genetic ancestry (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[2, 2], 2)*100`%) than children of mothers who did not (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[1, 2], 2)*100`%, *P* value=<0.001). Children of mothers who reported smoking during pregnancy had higher prenatal maternal smoking polymethylation scores than children of mothers who did not at both the age 9 (`r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) and age 15 (`r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) visits (Table 1, Supplemental Table 2). At age 9, children exposed to prenatal maternal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)`%) than children of non-smoking mothers (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`%, *P* value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`). At age 15, children exposed to prenatal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)`%) than children of non-smoking mothers, although this difference was not significant (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`% *P* value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).

@@

The association between prenatal maternal smoking and the polymethylation score was also observed in multivariable models, adjusting for base model covariates of child sex, maternal income to poverty ratio at baseline, proportion of salivary immune cells, sample plate from DNA methlation analysis, and child age  (Figure 3; Supplemental Table 2). At age 9, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score.  At age 15, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score. A consistent association was observed when stratifying to the African genetic ancestry sample (n= `r modeldata_people%>%filter(ancestry=='African ancestry')%>%nrow()`). In the African genetic ancestry sample at age 9, prenatal maternal smoking was associated with a `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score. The direction and magnitude of the association was similar in the European and Hispanic genetic ancestry samples at age 9, though attenuated at age 15 (Figure 3; Supplemental Tables 2).

At age 9, prenatal maternal smoking was associated with `r round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95% CI: `r paste(round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`) after adjusting forbase model covariates. Similarly, at age 15 prenatal maternal smoking was associated with `r round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95%CI: (`r paste(round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`)). The magnitude of the association between prenatal maternal smoking and cg05575921 was reduced in European genetic ancestry samples at age 15 (Figure 3). The direction of the association between prenatal maternal smoking and cg05575921 was reversed in Hispanic genetic ancestry sample (Figure 3).

Prenatal maternal smoking was not associated with global DNA methylation or epigenetic age (Figure 3).

## Sensitivity analyses for association testing

After additionally adjusting for other prenatal exposures and postnatal smoke exposure, the association between prenatal maternal smoking and the polymethylation score was very consistent (Supplemental Figure 4). Similarly, after additionally adjusting for other prenatal exposures and postnatal smoke exposure, the direction of the association between prenatal maternal smoking and cg055975921 was consistent, although no longer significant at age 15. Children exposed to postnatal smoke but not prenatal smoke did not have higher polymethylation scores than children unexposed to both postnatal and prenatal smoke.  (Supplemental Figure 5). Results were also similar when controlling for surrogate variables instead of known covariates (Supplemental Figure 6; Supplemental Table 2).

Results were similar when using linear mixed effect models instead of age-stratified models and adjusting for base model covariates (Supplemental Table 3 & 4).

Results were also similar when using polymethylation scores constructed using alternative regression coefficients (see Methods, Supplemental Figure 7). The association between prenatal maternal smoking and the polymethylation score using regression coefficients from newborn cord blood with cell-type controls was the strongest.  Prenatal maternal smoking was also associated with other selected *a priori* CpG sites consistently with associations reported in previous meta-analyses (Supplemental Table 2, Supplemental Figure 8).

@@

Next, we compared the accuracy of different DNA methylation summary measures for classifying prenatal maternal smoking using receiver operating curves (Figure 4A; Supplemental Table 5). We considered classification when using DNA methylation summary measures alone. We also considered classification when using DNA methylation summary measures in addition to a base model of child sex, maternal income to poverty ratio at baseline, child age at DNA methylation measurement, estimated immune cell proportion, plate from methylation processing and the first two components of genetic ancestry. When used without DNA methylation measures, these base model variables had an area under the curve (AUC) of `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 9')%>%pull(auc)` at age 9 and `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 15')%>%pull(auc)` at age 15.

At age 9, including the polymethylation score  significantly improved the base model accuracy (AUC: `r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(auc)`, *P* value comparing to base model<0.001).  At age 15, similarly, including the polymethylation score significantly improved the base model accuracy (AUC:`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(auc)` , *P* value<0.001).

At age 9, the base model with the polymethylation score had a larger AUC than the base model with cg05575921 (AUC base model + cg05575921: `r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(auc)`; *P* value =`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(5)`). At age 15, this was also true (AUC for base model + cg05575921=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(auc)`,  *P* value=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(5)`). Accuracy was not improved by including polymethylation scores from age 9 and age 15 together (Figure 4B).

## Sensitivity analyses for biomarker accuracy assessment

### 2021-11-10 Papers/Polymethylation_scores/06_discussion.Rmd

Our findings are also consistent with previous work on DNA methylation as a biomarker for prenatal maternal smoke exposure. Prenatal maternal smoking classification using saliva DNA methylation in our sample of 9 and 15 year olds performed comparably to classification using peripheral blood from older adults (AUC 0.72) [@Richmond:2018tu]. However, previous biomarkers of prenatal maternal smoking using DNA methylation of cord blood from newborns (AUCs ranging from 0.82-0.97) and peripheral blood from 3-5 year old children (AUC=0.87) more accurately classified prenatal maternal smoke exposure [@Reese:2017ts; @Ladd-Acosta:2016aa]. Differences in DNA methylation patterns across tissues and over time may influence the performance of DNA methylation biomarkers. The similarity in performance between saliva DNA methylation in our study and peripheral blood DNA methylation in older adults is encouraging for the use of saliva DNA methylation as a readily accessible clinical sample.

>>>>

Our findings are also consistent with previous work on DNA methylation as a biomarker for prenatal maternal smoke exposure. Prenatal maternal smoking classification using saliva DNA methylation in our sample of 9 and 15 year olds performed comparably to classification using peripheral blood from older adults (AUC 0.72) [@Richmond:2018tu]. However, previous prenatal maternal smoke exposure biomarkers created using LASSO regression and cord blood from newborns performed better (AUCs ranging from 0.82-0.97) [@Ladd-Acosta:2016aa] A support vector machine approach performed on and peripheral blood from 3-5 year old children also more accurately classified prenatal maternal smoke exposure (AUC=0.87) [@Reese:2017ts]. Differences in DNA methylation patterns across tissues and over time may influence the performance of DNA methylation biomarkers. Different methods for summarizing across multiple DNA methylation sites may also influence biomarker performance. The similarity in performance between saliva DNA methylation in our study and peripheral blood DNA methylation in older adults is encouraging for the use of saliva DNA methylation as a readily accessible clinical sample.

### 2021-11-10 Papers/Polymethylation_scores/99_FigureCaptions.Rmd

#### Figure 1 - Selection of samples from the Fragile Families and Child Wellbeing study into analytic subset. N represents the number of individuals at each step in the selection procedure, M represents the number of samples. Individuals with repeated measures can have more than one sample.

@@

#### Figure 2 - Differences in selected DNA methylation summary measures by self-report of prenatal maternal smoking among `r n_total` children in the Fragile Families and Child Wellbeing study (CT=cell-type control)

@@

#### Figure 3 - Prental maternal smoke exposure is consistently associated with polymethylation scores at ages 9 and 15 and is portable across genetic ancestry groups in a sample of `r n_total` children in the Fragile Families and Child Wellbeing study

@@

#### Figure 4 - A) Polymethylation scores accurately classify maternal report of prenatal smoke exposure status at ages 9 and 15 among a `r n_total` children in the Fragile Families and Child Wellbeing Study B) Inclusion of two-visits of DNA methylation data does not improve classification of maternal report of prenatal smoke exposure status among `r n_total` children in the Fragile Families and Child Wellbeing study

>>>>

#### **Figure 1 - Selection of samples from the Fragile Families and Child Wellbeing study into analytic subset.** N represents the number of individuals at each step in the selection procedure, M represents the number of samples. Individuals with repeated measures can have more than one sample.

@@

#### **Figure 2 - Differences in selected DNA methylation summary measures by self-report of prenatal maternal smoking among `r n_total` children in the Fragile Families and Child Wellbeing study at ages 9 and 15.** Samples from children exposed to prenatal maternal smoke in grey, samples from children unexposed to prenatal maternal smoke in black. From top-left, clockwise: Polymethylation scores for prenatal maternal smoke exposure, constructed using regression coefficients for prenatal smoke exposure predicting DNA methylation in newborn cordblood samples, accounting for cell-type control. DNA methylation values from samples in the Fragile Families and ChildWellbeing study were mean-centered, then multiplied by these regression coefficients and summed. Pediatric epigentic clock (years). AHRR gene: percent cg05575921 methylation. Percent global DNA methylation.

@@

#### **Figure 3 - Prenatal maternal smoke exposure is consistently associated with polymethylation scores at ages 9 and 15 and is portable across genetic ancestry groups in a sample of `r n_total` children in the Fragile Families and Child Wellbeing study** All models shown controlled for: first two principal components of child genetic ancestry (from ancestry-stratified principal components for ancestry stratified models), child sex, maternal income-to-poverty ratio at birth, immune cell proportion estimated from methylation data, yes/no other maternal prenatal drug use, yes/no maternal prenatal alcohol use, postnatal maternal smoking when child age 1 or age 5, postnatal maternal/primary care give smoking packs/day in month prior to saliva sample.

@@

####  **Figure 4 - Polymethylation scores accurately classify prenatal maternal smok exposure at ages 9 and 15 among `r n_total` children in the Fragile Families and Child Wellbeing study** A) Receiver operator curve for select DNA methylation measures for predicting prenatal smoke exposure using no other variables (light colors) or using base model variables (dark colors, other variables included: child sex, maternal income-poverty ratio at birth, immune cell proportion and batch of methylation data processing). B) Receiver operator curve for include polymethylation scores individually at each visit (black & light grey) or jointly (dark grey).
